Data-Driven Modelling \& \\ Forecasting of Chaotic Systems

### 2022-01-17 draft_16Jan/hiscurrent.tex
Grammarly changes copied in

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,[?, ?]) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice, provide poor long-term consistency for a large class of systems. (Citation)

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding [?] when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. Takens delay embedding theorem [?] and its various generalisations (e.g., [?, ?, ?]) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (e.g., [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter 3), it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique, and consequently, one wishes that these would have low functional complexity[?], i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks (Citations to echo state networks)) by mapping data onto a higher dimensional space for further processing. Although they perform well on forecasting some dynamical systems, they fail on others (Citations here) as there is often no guarantee that the correct function was learnt during training.

This thesis involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method ([?]) that incorporates learning a function by mapping the data onto a higher dimensional space using what is called a driven dynamical system (See Chapter 4). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows:
\newline In Chapter 2, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE, and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems.
\newline Chapter 3 introduces the problem of forecasting dynamical systems, states the Takens delay embedding theorem, and discusses various issues faced while forecasting.
\newline In Chapter 4, a driven dynamical system is defined and discusses the properties of a specific class of driven dynamical systems that we use in this thesis/report/paper.
\newline Finally, Chapter 5 shows the implementation of these forecasting methods, and conclusions are provided in Chapter 6.

@@

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this thesis, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model, in this case, is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

Formally, a function $T : U \to U$, where U is some set is a discrete-time dynamical system and its iterates ${u, Tu, T^{2}{u},\ldots}$, where Tn denotes the n-fold composition of T with itself, describe the evolution of an initial condition $u \in U$ (Note that we frequently drop the brackets and represent $T(u)$ by $Tu$ to simplify notations).

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed. Citation 
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to \cite{saber2010introduction}.

@@

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However, when the space U is compact, all orbits approach an invariant set. It follows because one can show the set of limit points of the orbit to be invariant~\cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

@@

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples Tu = u/2 and Tu = 2u defined on R (where {0} “attracts” orbits) or the example, Tu = u2 defined on [0,1] (where {1} “repels” orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular, we are interested in those invariant sets named attractors.

@@

Indeed in our previous examples, it can be verified that the system  $Tu=u/2$ defined on  $\mathbb{R}$ and $Tu=u^2$ defined on $[0,1]$  the singleton set $\{0\}$ is an attractor.  For the system,  $Tu=1-|2u-1|$ on $[0,1]$, one may easily verify that the only attractor is the entire space $[0,1]$. This follows from the fact that between any two points $u< v$ in $[0,1]$ and for any $a,b$ so that $u\le a < b \le v$, we can find an $n$ so that $T^n(a,b)=[0,1]$ (as would be explained later in this chapter).

A dynamical system can have several attractors and may also be contained in another attractor. For the example, $Tu = ^u2$ on $[0,1]$, both $\{0\}$ and [0,1] are attractors. It is known that an attractor for a dynamical system always exists in a compact space. \textbf{(Cite).}

The dynamics restricted to an invariant set can be complicated. For instance, an invariant set could be just a single point, or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible. A particular well-studied phenomenon of such complexity gives rise to so-called chaotic behaviour, studied in detail over the past fifty years.

@@

In the 1960s, several mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper [?]. The notions of invariance, attractivity, and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. The narrative of Lorenz's discovery of chaos and the history of other forerunners in this subject is fascinating. We highly recommend James Gleick's book Chaos: The Making of a New Science [?] for those interested. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery and describes the underlying mathematical notions for non-specialists.

Several intricately different definitions of chaos are available in the literature (Cite), each indicating some aspect of complexity. In practice, it is not possible to verify which definition or notion the underlying dynamical system satisfies when only data is observed from it. However, we specifically recall the definition of chaos in the sense of Devaney [?, ?] to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion of sensitive dependence on initial conditions.

@@

It is vital to acquire a sense of this definition as it can easily be misunderstood. It is a common misconception to interpret SDIC as two close points (u and v) that eventually become separated by a distance $\delta$ under iteration by T, but this is not true. To fully comprehend all the subtleties of the concept, one needs to discuss it more thoroughly by considering each sentence with care and attention; whereafter, one may examine how they fit together to convey the idea of sensitive dependence.

@@

Topological transitivity implies that iterates of an open set of initial conditions get mixed up with other open sets. On a compact metric space, one may show that topological transitivity also implies the existence of a point whose forward iterates are dense [?]; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one whose iterates are almost dense.

@@

To be more precise (see \cite{de2013elements}), for a discrete-time dynamical system on a compact space, the set of points with dense iterates are residual, and they are typical or likely to be observed in practice. In this thesis, when data is observed from a topologically transitive system, we assume it arises from a dense orbit.

@@

%Composite functions
To understand the diagram, we note that if we travel "right and then down," the diagram instructs us to use T (top arrow right) first, followed by $\phi$. (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed "down and then right", the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

@@

A practical example of this would be the subsequent scenario: given the time-sequential coordinates of an object moving in space, predict the future positions of that object. However, we are very rarely, if at all, presented with a problem where the entire state information is available to us in the form un at some timestep $n\in\mathbb{N}$. Consider then a more complicated learning problem:

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e.\ednote{two i.e.'s}, we examine  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that, though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. In almost all cases, the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{M: Maybe replace difficulty by saying that we need to consider two aspects. B: Perhaps this edit makes more sense?}
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

\section{Practical Limitations to Takens' Theorem}

Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

@@

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of another requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when T maps the space U into a proper subset B of U, there exists some point in $U - A$ that does not have a preimage and we conclude that T is not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, the dying oscillations for a damped pendulum. (A more in-depth account will be provided in 5)

@@

In this chapter, we discuss results on the mapping of temporal data obtained from a discrete-dynamical system onto a different space through the notion of a driven dynamical system. 
We also consider the conditions a driven system should possess to avoid adding distortion to its state-space representation. We then describe the single-delay Dynamics (SDD) of the system and finally recall conditions on the driven system so that the SDD are conjugate (or at least semi-conjugate) to the underlying system. 
The SDD can then be used to forecast and reconstruct the underlying system.

@@

To find a function $\phi$, the need  now arises to consider a so-called driven dynamical system, a special case of a nonautonomous dynamical system. Two spaces are considered: input is taken from an input space U and the state of a space X is updated. (Previously, only the state was considered.)
We do this to mimic the true conditions. By accounting for an extrogenous input sequence in $U$ which can also influence the  system-state at a specific time-step, it produces more general models than an autonomous system.

@@

The update equation $x_{n+1} = g(u_n,x_n)$ for $n \in\mathbb{Z}$, input $u_n$ from $U$ and state $x_n$ belonging to $X$ (a compact space) generate the dynamics on X. 
Abbreviated, we refer only to the \textit{driven system g} where all other entities are understood implicitly.

@@

As the solutions to a driven system are often considered an important entity, we next identify a subspace $X_U$ of $X$ that contains all possible solutions. To realise such a subspace of a driven system $g$, the concept of a reachable set is defined.

@@

One of the first results obtained after defining the USP is that every solution will attract different initial conditions towards the component parts of the solution. 
If $g$ has the USP, then any solution to $g$ is also a uniform attractor in nonautonomous dynamical systems literature \cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this project, and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. 
\ednote{B: edited here}

@@

SI-invertibility promises that 'enough' information is retained when $g$ is chosen without introducing one's choice of $g$ without introducing unwanted complexity. 
'Enough' here refers to the fact that we may always recover the previous input value $u_n$ (if we know successive states $x_{n-1}, x_n$).

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: So that FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: Note! We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

Our implementation of the reservoir(A) and input(B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee a measure of stability. /ednote{B: Not satisfactory. Edit again.}

>>>>

Recent partially interpretable models available in the literature have been based on the Koopman operator (e.g.,[?, ?]) to employ observables mapping the data onto a higher-dimensional space. This makes the dynamics in the higher-dimensional space more amenable for approximation by a linear transformation. Such methods do not guarantee exact reconstructions for nonlinear models, and in practice, provide poor long-term consistency for a large class of systems. (Citation)
\textbf{(Citation)}

The non-interpretable models include the delay embedding and the machine learning algorithms. For example, one could learn a system conjugate to the underlying system by applying the Takens delay embedding~\cite{takens1981detecting} when one has "good" observations from the system; this learnt system could then be used to forecast the observed data. 
Takens delay embedding theorem ~\cite{takens1981detecting} and its various generalisations (e.g., \cite{sauer1991embedology, stark1999delay, gutman2018embedding}) establishes the learnability of a system constructed by concatenating sufficiently large previous time-series observations of a dynamical system into a vector (called delay coordinates). This then confirms the existence of a map on the space of delay coordinates equivalent (or topologically conjugate) to the underlying map from which the observed time-series was first obtained. Although topological conjugacy guarantees an alternate representation of the underlying system, the quality still depends on numerous parameters, making the comprehension of the dynamics unreliable (e.g., [?]). One reason for this fragility is that the embedded attractor in the reconstruction space is not always an attractor of the map learnt in the reconstruction space, despite unquestionably being an invariant set. When the embedded object is not explicitly known to be an attractor (as explained in Chapter~ \ref{ch3}), it can cause predictions to fail.

Practically, the application of Takens embedding involves learning a map through some technique, and consequently one wishes that these would have low functional complexity\cite{manjunath2021universal}, i.e., functions with fewer oscillatory graphs. On the other hand, pure machine learning methodology processes temporal information (like the echo state networks (Citations to echo state networks)) by mapping data onto a higher dimensional space for further processing. Although they perform  well on forecasting some dynamical systems, they fail completely on others (Citations here) as there is often no guarantee that the right function was learnt during training.

This thesis involves implementing and analysing non-interpretable models that can guarantee exact reconstruction. 
The project work concerns the study and implementation of a method (\cite{manjunath2021universal}) that incorporates learning a function by mapping the data on to a higher dimensional space using what is called a driven dynamical system (See Chapter~ref{ch.4}). With a clear understanding of how the data is mapped onto the higher dimensional space, the method then permits the learning of a dynamical system topologically conjugate to that of the underlying system. Instead of linear regression as in the training of echo state networks, deep learning methods are employed to learn the correct function. With slight modifications to the implementation in the paper (with Adriaan), we show that one can construct accurate non-interpretable models with the ability to reconstruct attractors from more hard-to-forecast systems like the double pendulum. (The forecasting of the time-series from a double pendulum has not been reported before.) Moreover, we also demonstrate that long-term statistical consistency is preserved.

@@

The more intricate details of proofs are referred to where relevant throughout. This report is organised as follows: 
\newline In Chapter~\ref{ch2}, we recall the definition of a discrete-time dynamical system, how a discrete-system arises from a flow of an ODE and then proceed to define the inverse-limit space and topological conjugacy of autonomous systems. 
\newline Chapter~\ref{ch3} introduces the problem of forecasting dynamical systems, states the Takens delay embedding theorem, and discusses various issues faced while forecasting. 
\newline In Chapter~\ref{ch4}, a driven dynamical system is defined and discuss the properties of a specific class of driven dynamical systems that we make use of in this thesis/report/paper. 
\newline Finally, in Chapter~\ref{ch5} we show the implementation of these forecasting methods, and conclusions are provided in Chapter~\ref{ch6}.

@@

At its most elementary level, a dynamical system is just something that evolves deterministically through time. In the context of this thesis, deterministic refers to the fact that a system evolves according to specified rules rather than based on random events. 
Dynamical systems arise in a variety of situations. A continuous-time dynamical system describes the states for all values of the time. Specifically, if the motion of a pendulum in which the quantities such as the angular position and angular momentum are known at all times, then it is a continuous-time dynamical system. The equations of the dynamical system can take the form of one or more ordinary differential equations that determine the relevant quantities at any future time if we know the initial location and momentum. 
In ecology, discrete-time dynamical systems are widely used to model population growth. The model in this case is a function calculating the following generation's population given the population of the previous generation. If we know the starting population, we may once again calculate the population at any time in the future.

Formally, a function $T: U \to U$, where $U$ is some set is  a \emph{discrete-time dynamical system} and its iterates $\{u,Tu,T^2u,\ldots\}$, where $T^n$ denotes the $n$-fold composition of $T$ with itself, describe the evolution of an initial condition $u\in U$ (Note that we frequently drop the brackets and denote $T(u)$ by $Tu$ so as to simplify notations).

@@

Of course, discrete-time dynamical systems need not always arise through a differential equation. Once again, we may consider the field of ecology, where discrete dynamical systems are often directly derived or assumed. \textbf{Citation}
There is a school of thought that advocates discrete-time dynamical systems to be more natural for modelling real-world observations than differential equations. We refer the interested reader to \cite{saber2010introduction}.

@@

Two examples of invariant sets include a fixed point where $Tu=u$ for  $u\in U$, and a periodic orbit, i.e., a set of iterates $\{u,Tu, T^2u,\ldots,T^pu\}$, where $T^{p+1}u=u$ for some $p\in\mathbb{Z}$.  The entire space $U$ could be also be invariant.
Consider for example the space $U=[0,1]$, where $Tu=4u(1-u)$ and then $U$ is invariant (as every $u\in{U}$ can be written as $u = 4x(1-x)$ for some $x\in{U}$).

@@

However,  when the space $U$ is compact, all orbits approach an invariant set. This is since the set of limit points of the orbit can be shown to be invariant \cite{de2013elements}. So, when $U$ is compact,  the $\omega$-limit set $\omega(u;T)$ of a point $u$ defined to the collection of limit points of the sequence $\{x,Tu,T^2u,\ldots\}$ is nonempty, and $\omega(u;T)$ is invariant.

@@

Invariant sets have various properties. Invariant sets can be attracting or repelling depending on how orbits in their vicinity behave. 
Recall the examples $Tu =u/2$ and $Tu=2u$ defined on $\mathbb{R}$ (where $\{0\}$ “attracts" orbits) or the example, $Tu=u^2$ defined on $[0,1]$ (where $\{1\}$ “repels" orbits). 
We are interested in attractive invariant sets since they capture the long-term dynamics as time increases. 
In particular, we are interested in those invariant sets named attractors.

@@

Indeed in our previous examples, it can be verified that the system  $Tu=u/2$ defined on  $\mathbb{R}$ and $Tu=u^2$ defined on $[0,1]$  the singleton set $\{0\}$ is an attractor.  For the system,  $Tu=1-|2u-1|$ on $[0,1]$, one may easily verify that the only attractor is the entire space $[0,1]$. This follows from the fact that between any two points $u< v$ in $[0,1]$, and for any $a,b$ so that $u\le a < b \le v$, we can find an $n$ so that $T^n(a,b)=[0,1]$ (as would be explained later in this chapter).

A dynamical system can have several attractors and may also be contained in another attractor. For the example, $Tu =u^2$ on $[0,1]$, both $\{0\}$ and $[0,1]$ are attractors. It is known that an attractor for a dynamical system always exists in a compact space. \textbf{(Cite).}

The dynamics restricted to an invariant set can be complicated. For instance an invariant set could be just a single point or it could have an infinite set. If the invariant set is infinite, then complicated dynamics are possible.A particular well-studied phenomenon of such complexity gives rise to so-called  chaotic behaviour, studied in detail over the past fifty years.

@@

In the 1960s, several mathematicians and mathematically interested scientists independently discovered chaos in the mathematical sense. The meteorologist Edward Lorenz may have been the first to explain this phenomenon in his 1963 paper \cite{lorenz1963deterministic}. The notions of invariance, attractivity, and chaos may also be described for continuous systems, and Lorenz's system comprised of a system of differential equations. 
The narrative of Lorenz's discovery of chaos and the history of other forerunners in this subject is fascinating. We highly recommend James Gleick's book Chaos: The Making of a New Science~\cite{gleick2008chaos} for those interested. It clearly illustrates these experiences, explains why chaos was such a startling and crucial mathematical and scientific discovery and describes the underlying mathematical notions for non-specialists.

There are several intricately different definitions of chaos available in the literature \textbf{(Cite)}, with each of them indicating some aspect of complexity.
In practice, it is not possible to verify which definition or notion does the underlying dynamical system satisfies when only data is observed from it. We do, however, specifically recall the definition of chaos in the sense of Devaney \cite{devaney2018introduction,de2013elements} so as to understand some nuances behind the complexity. Devaney's definition of chaos has three requirements, with the first being the notion  of sensitive dependence on initial conditions.

@@

It is vital to acquire a sense of this definition as it can easily be misunderstood. It is a common misconception to interpret SDIC as two close points ($u$ and $v$) that eventually become separated by a distance $\delta$ under iteration by $T$. But this is not true. To fully comprehend all the subtleties of the concept, one needs to discuss it more thoroughly by considering each sentence with care and attention, whereafter one may examine how they fit together to convey the idea of sensitive dependence.

@@

Topological transitivity implies that iterates of an open set of initial conditions get mixed up with other open sets. On a compact metric space, one may show that topological transitivity also implies the existence of a point whose forward iterates are dense \cite{de2013elements}; or in other words, the orbit going through this point will be dense in the compact metric space. 
In fact, it is topologically more likely that the choice of an arbitrary point will be one whose iterates are almost dense.

@@

To be more precise (see \cite{de2013elements}), for a discrete-time dynamical system on a compact space, the set of points with dense iterates are residual, and they are typical or likely to be observed in practice. In this project, when data is observed from a topologically transitive system, we assume it arises from a dense orbit.

@@

%Composite functions
To understand the diagram, we note that if we travel "right and then down," the diagram instructs us to use $T$ (top arrow right) first, followed by $\phi$. (right arrow downwards). Consequently, the pathway amounts to finding $\phi(T(u))$. If we proceed "down and then right," the diagram instructs us to apply $\phi$ (left arrow downwards) first, and then apply $S$ (bottom arrow right). This then amounts to finding  $S(\phi(u))$. When the relationships in this diagram hold, we say $\phi(T(u))= S(\phi(u))$, and we formally denote it as $\phi \circ T=S\circ \phi$.

@@

A practical example of this would be the subsequent scenario: given the time-sequential coordinates of an object moving in space, predict the future positions of that object. We are very rarely, if at all, presented with a problem where the entire state information is available to us in the form $u_n$ at some timestep $n\in\mathbb{N}$. Consider then a more complicated learning problem:

@@

First we establish the method of Takens delay embedding. 
In this method, one considers a discrete-time dynamical system defined on a 'nice' space (a smooth manifold)\ednote{B: Should this be further explained?} that can be obtained as time-$K$ map of a continuous time-dynamical system, concepts formalised below.
We make observations from such a system, i.e., we observe the evolution of $\theta(w_0)$ (where $w_0$ represents the initial state) i.e.\ednote{two i.e.'s}, we examine  a finite set of values of $u_n :=  \theta(w_n) = \theta(Tw_{n-1})$. The sequence $\{u_n\}$ represents a scalar time-series and intuitively $\theta$ represents a probe inserted into a bigger system which is itself only measuring/extracting a small part of the greater system state at time $t=n$. 
Consider, for example, a thermometer erected to measure the ambient temperature in a local village. This measurement function, the thermometer, is capturing only a single aspect, the temperature, of a much grander dynamical system entailing the current weather of the surrounding area. Even more than that though- it is measuring a miniscule part of the global weather system.

@@

However, the actual state $u$ of a system is seldom, if ever, fully known. the most one can do is to insert probes into a system to obtain partial information through the measurements taken. Moreover, the process of taking a measurement itself introduces two additional aspects which complicate the problem: \ednote{B: Perhaps this edit makes more sense? M: Maybe replace difficulty by saying that we need to consider two aspects. }
\vspace{-8mm}
\begin{enumerate}[noitemsep, label=\roman*.]
  \item A series of measurements over a specific time interval is inherently a discretisation procedure of the underlying continuous-time dynamical system. (Hence why we restricted our attention to discrete-time systems in the preceding section)
  \item A probe will never be entirely accurate, and so the act of measurement introduces a certain measure of numerical noise/inaccuracy.
\end{enumerate}

@@

\section{Practical Limitations to Takens' Theorem}
Takens' Embedding Theorem is a powerful result, and it provides compelling reason to believe that one could conceivably reconstruct a system conjugate accurately to the underlying system. 
Nonetheless, it does present some severe practical limitations:
\vspace{-5mm}
\begin{enumerate}
\item Even supposing that we could indeed find $F_\theta$ , our \emph{approximation} of  $F_\theta$ is a map from a larger set $\mathbb{R}^{2d+1}$ containing the embedded attractor. There are, however, no theoretical guarantees that $F_\theta$ will retain  $V=\Phi_{2d,\theta}(W)$  as an attractor although $W$ could itself be an attractor.
\item Takens theorem is stated only for noiseless observations. Due to noise $\epsilon_n$ the delay vector $\Phi_{2d,\theta}(w_n) + \epsilon_n$  may lie outside $V$.Furthermore, due to the chaotic nature of the underlying system (i.e. the fact that it has SIDC), the evolution of $\Phi_{2d,\theta}(w_n) + \epsilon_n$ under the map $F_\theta$ could move out of $V$ completely. This problem can be overcome  by using a driven dynamical system with some properties, and we discuss this in Chapter\ref{ch3}. 
\end{enumerate}

@@

Though not a uniquely related to the Takens embedding theorem, we do opt to make mention of another requirement for our work. 
The methods described here are applicable for data originating from a surjective map. 
This follows from the fact that when $T$ maps the space $U$ into a proper subset $B$ of$ $U, there exists some point in $U - A$ that does not have a preimage and we conclude that $T$ is not surjective. 
An example of this would be the case when a system displays energy-loss. Consider, for instance, consider the dying oscillations for a damped pendulum. (An account of greater depth  is provided in~\ref{ch5})

@@

In this chapter, we discuss results on the mapping of temporal data obtained from a discrete-dynamical system onto a different space through the notion of a driven dynamical system. 
We also consider the conditions a driven system should possess to avoid adding distortion to its state-space representation. We then describe the \emph{single-delay dynamics} (SDD) of the system and finally recall conditions on the driven system so that the SDD are conjugate (or at least semi-conjugate) to the underlying system. 
The SDD can then be used to forecast and reconstruct the underlying system.

@@

To find a function $\phi$, the need now arises to consider a so-called driven dynamical system, a special case of a nonautonomous dynamical system. Two spaces are considered: input is taken from an input space $U$ and the state of a space $X$ is updated. (Previously, only the state was considered.)
We do this to mimic the true conditions. By accounting for an extrogenous input sequence in $U$ which can also influence the  system-state at a specific time-step, it produces more general models than an autonomous system.

@@

The update equation $x_{n+1} = g(u_n,x_n)$ for $n \in\mathbb{Z}$, input $u_n$ from $U$ and state $x_n$ belonging to $X$ (a compact space) generate the dynamics on X. 
Abbreviated, we refer only to the \textit{driven system $g$}, where all other entities are understood implicitly.

@@

As the solutions to a driven system are considered an important entity, we next identify a subspace $X_U$ of $X$ that contains all possible solutions. To realize such a subspace of a driven system $g$, the concept of a reachable set is defined.

@@

One of the first results obtained after defining the USP is that every solution will attract different initial conditions towards the component parts of the solution.
If $g$ has the USP, then any solution to $g$ is also a uniform attractor in nonautonomous dynamical systems literature~\cite{Manju_Nonlinearity}. The discussion on nonautonomous attractors is beyond the scope of this project, and we refer the reader to \cite{Manju_ESP, esann2012ids} for further reading.

@@

The driven system $g$ can induce an embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ as follows: 
If the causal mapping $H:\overleftarrow{U}{\to}{\overleftarrow{X}_U}$ is injective (in addition to being surjective), it becomes the embedding of $\overleftarrow{U}$ in $\overleftarrow{X}$ induced by the driven system $g$. 
The continuity of $H^{-1}$ follows from the fact that $H$ is itself a continuous and surjective mapping of a compact space $\overleftarrow{U}$ in a Hausdorff space.
and we refer to the map $H$ as a \emph{causal embedding}. 
\ednote{B: edited the sentence here}

@@

SI-invertibility promises that 'enough' information is retained when $g$ is chosen without introducing one's choice of $g$ will still ensure 'enough' information is retained without introducing unwanted complexity. 
'Enough' here refers to the fact that we may always recover the previous input value $u_n$ (if we know successive states $x_{n-1}, x_n$).

@@

We implement a feedforward network, (a network with no memory, unlike the RNN)\ednote{B: Added so that FFN is in more context} required to learn $\Gamma$ in the \emph{Python} programming language and making use of the \emph{Scipy} library, alongside the \emph{Keras} package in the \emph{Tensorflow} library \ednote{B: Is this something you want me to write in? M: Note! We do not implement the RNN, but the FN that learns Gamma through the Tensorflow. We called the RNN + FN set up a RCN in Adriaan's paper. FYI: FN's work only if there is a definite relationship between the $u_n$ and $u_{n+1}$ which is rare in temporal data.
RNNs with ESP work when there is a definite relationship between $(\ldots,u_{n-1},u_n)$ and $u_{n+1}$ and is suited for temporal processing.}.

@@

Our implementation of the reservoir(A) and input(B) matrices in the RNN makes use of sparse matrices. We discuss three specifics relating to this concept. Firstly, the matrices are assigned a variable density in terms of the amount of non-zero entries and are scaled to have values within the range $[-1,1]$. For this, we make use of \textit{Python's SciPy} library and specifically the \textit{Sparse, LinAlg} modules. In our experiments, we generally opted for a density ratio of 10\%-25\%.
Secondly, we opted to enforce the requirement that both matrices be well-conditioned, i.e. that they have a condition number of \~1. The condition number of a matrix $M$ (denoted $cond(M)$) is the product of the matrix norms of the matrix $M$ and its inverse $M^{-1}$ and is a measure of the sensitivity of output of a matrix multiplication when an input is changed. \textbf{(Cite.)} By ensuring a condition number close to 1, we also ensure that our driven function $g$, depending on the matrices $A$ and $B$, will not be sensitive to adjustments in the input. In so doing we then also guarantee a measure of stability. \ednote{B: Not satisfactory. Edit again.}
